---
title: "Examining the case behind Acer Platanoide's plight on Toronto's Street Trees "
subtitle: ""
author: 
  - Veyasan Ragulan
thanks: "Code and data are available at: https://github.com/Veyasan1/STA3014-Paper1"
date: today
date-format: long
abstract: "Acer platanoides is an invasive species that has taken over the streets of Toronto. Using the Street Tree data from OpenData Toronto, this paper examines the patterns and trends behind this spread. A linear model was created to predict species based on ward, street and cross streets. The model didn't find directionalyty to Acer platanoides' spread, suggested that volumes of Acer platanoides should be monitored instead."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(arrow)
library(modelsummary)
library(kableExtra)

analysis_data <- read_csv(here("data/analysis_data/analysis_data.csv"))
model_data <- read_parquet(here::here("data/model_data/model_data.parquet"))
model <- readRDS(here("models/model.rds"))
```

# Introduction

Urban spaces can be void of vast expanses of nature, whether it be grass fields, blooming flowers, or teaming and diverse natural wildlife. One simple way to tackle this is to incorporate nature within the urban landscape, such as with trees. Trees are a vital plant in the world, they take in carbon dioxide and provide oxygen, a key component in the air we breath. Urban areas are known for the excess in carbon dioxide, which only makes it seem natural that trees can live and perhaps thrive in such environments. One way cities have incorporated trees into the city are roadside trees. Trees line many of the cities highways, arterial roads, collector roads, and small roads. Lines of trees lined alongside highways can act as a natural sound barrier by reflecting, scattering and absorbing nearby noises [@aais]. Tree-lined streets can also alter the way wind affects pedestrians below them, creating a more hospitable environment to promote active transportation such as walking or biking [@beptree]. It should be no surprise then, that many cities around the world, including Toronto, have incorporated trees into their landscape, in busy downtown centres, to sleepy suburbs, and everywhere in between.

Acer platanoides is a well known tree used in urban environments, especially after the demise of elm trees in the 1970s [@urbanforest]. These trees are know for ‘a vigorous early growth rate, desirable form and size, the capacity to withstand many urban impacts (e.g. pavement, moderate levels of pollution, dusts, and dry soils) and the abilities to transplant well, grow on a wide variety of soils, and withstand ice and snow damage better than other maples’ [@norwaymaple]. Being planted in cities, it's seeds can travel by wind across to more wooded suburbs, where the seeds can tolerate shade for extended periods of time [@urbanforest]. However, as it's common name suggests, it is not a native species to Canada, instead brought here by European explorers and settlers. While not a very invasive species, taking over just 9% of Southern Ontario's forests, it does so by harming other native plants and trees [@urbanforests]. More attention should be cast on the spread of this tree, as it has the potential to devastate Toronto's and Southern Ontario's forests in the years to come.

In this paper we look at street tree data from Toronto's roadside trees, and discern any patterns or observations behind the rise of Acer Platanoides. To do this, a linear model was created to predict botanical_name. Predictors include ward, streetname, and crosstreet1 and crossstreet2. The linear model failed to find a specific direction Acer platanoides travels in, rather it is present everywhere. This suggested that instead of looking for directions Acer platanoides spreads in, we should focus on the counts of Acer platanoides in a given area.

Section 2 will outline the source of this data, and highlight key varaibles to be used in the model. Section 3 covers the construction and rationale of the model. Section 4 is where the model results will be outlined. Finally, section 5 discusses key takeaways from the data and the model, and addresses weaknesses and limitations that can be considered for another report.

# Data {#sec-data}

## Measurement

All collection and analysis of data was done using R, a free programming language designed for data scientists [@citeR]. Additionally, tidycerse [@citeTidyverse], dplyr [@citeDplyr], and janitor [@citeJanitor] have been used as packages to read, clean, and present data. Finally OpenData Toronto is an online database containing thousands of datasets pertaining to the City of Toronto, which is where the data for this paper was initially accessed [@citeOpendatatoronto].

Rohan Alexander's book, *Telling Stories with Data* was referenced for troubleshooting and general ideas for data analysis [@tellingstories]

The data comes from Urban Forestry, an organization dedicated to Toronto's urban forests. They work towards planting more trees, and protecting existing trees from damage due to individuals, private entities, and public entities [@urbanforestry]. This particular dataset is called Street Tree Data, and focuses on city-owned trees located on roads. This information would be most likely used by city planners and road maintenance, ensuring snow plows and street dusters don't impede on any trees that line the roads. It is important to note that there may be some privately owned trees listed among the municipal trees, as they may be of interest with regards to road maintenance or other city services.

Street Tree Data contains around 32000 observations, each observation being an individual tree. Each observation has a general id for data analysis, as well as an structure id, telling us if the tree is part of an existing structure or building. Location is covered in 4 ways: first the address number (address) and streetname (streetname) of nearest postal delivery address, next is the street name and the nearest 2 cross streets (crosstreet1 and crossstreet2), third being the ward number (ward), and finally geometry contains a tuple of general latitude and longitude. Street Tree contains both the species name (botanical_name) and the common English name common_name), however we have decided to use only the species name to avoid confusion between similar sounding trees. Common English names will be provided for species of key interest. There is also measurement called dbh_trunk. This is the standard method forestry experts use to measure the diameter of a tree. Depending on where the measurement is taking place, the breast height changes. For example, in Canada it is 1.3 meters, while the US uses 4.5 feet. This methodology is comprised of old conventions used to measure trees used throughout the world [@dbh]. The result is a complicated mess where breast_height is a vague term that changes by location. Additionally, the convention breaks down when the tree is anything but vertically upright and straight. As a result, the DBH is more of an estimate rather than a sturdy measurement.

## `botanical_name`

To see why Acer platanoides is of interest in this paper, refer to @fig-1. The most planted street tree in Toronto is Acer platanoides, which has at least a 2x lead over any other tree in the data. This trend doesn't seem to be coincidental, and from prelimiary research, the prevalance might be worrisome. Therefore, the paper will use `botanical_name` as the target variable in a linear model, which can be filtered down to Acer platanoides later.

```{r}
#| label: fig-1
#| fig-cap: 10 most populous street tree species in Toronto, Ontario
#| echo: false
#| warning: false
#| message: false

species_count <- analysis_data %>%
  count(botanical_name, sort = TRUE) %>%
  head(10)

species_count |> 
  ggplot(aes(y = botanical_name, x = n)) +
  geom_col() +
  theme_minimal() +
  labs(x = "Planted Trees",
       y = NULL)
```

## `ward`
Ward is one of several location varaibles included in the Street Trees dataset. It lists the ward location of each observation (tree) in the city. Toronto is broken up into municipal wards, who is represented in City Hall by a Councillor. Wards vary in size and population, which may have an impact in the spread of Acer platanoides. For example, one of the largest Wards is Etobicoke North, which forms the north-western boundary of Toronto. Given its size and distance from downtown Toronto, it has ample room to host large swaths of forests, and have its roads lined with trees, increasing the possibility of Acer platanoides finding its way here. Toronto-Centre is the smallest ward by far, and being situated in downtown Toronto, is more dense with human infrastructure. There is less space for Acer platanoides to grow here, but the few green spots here could be susceptible to Acer platanoides.

```{r}
#| label: fig-2
#| fig-cap: Acer platanoides distribution across Toronto wards up to September 2024
#| echo: false
#| warning: false
#| message: false

# Gather counts of Acer platanoides per ward
ward_count <- analysis_data %>%
  filter(botanical_name == "Acer platanoides") %>%
  count(ward, sort = TRUE)

ward_count |> 
  ggplot(aes(x = ward, y = n)) +
  geom_col() +
  theme_minimal() +
  labs(x = "Ward",
       y = "Acer platanoides count")
```
According to @fig-2, Etobicoke North (Ward 2) hosts the most trees of the species Acer platanoides, and Toronto Centre (Ward 13) hosts the least. This falls in line with what was said before.

## `streetname`
Anoter location variable in the Street Trees dataset, `streetname` records the street each observation(tree) lies on. Streets come in many sizes and lengths across Toronto. Arterial streets such as Bloor st / Danforth Ave might be able to hold more trees due to their length, but given their capacity, raod designers may have opted to not leave space for tree growth alon certain stretches. On the other end of the spectrum, small roads such as Eireann Quay, will have less street length for trees, but they may be quiet roads which prioritze natural elements such as trees.

```{r}
#| label: fig-3
#| fig-cap: Acer platanoides distribution across Toronto streets up to September 2024
#| echo: false
#| warning: false
#| message: false

# Gather 10 streets with the highest counts of Acer platanoides
street_count <- analysis_data %>%
  filter(botanical_name == "Acer platanoides") %>%
  count(streetname, sort = TRUE) %>%
  head(10)

street_count |> 
  ggplot(aes(x = n, y = streetname)) +
  geom_col() +
  theme_minimal() +
  labs(x = "Street",
       y = "Acer platanoides count")
```
@fig-3 shows that Keele street holds the most amount of Acer platanoides in Toronto. Keele St is a North-South arterial road that starts at the north-eastern tip of High Park. This suggests that proximity to large parkscapes such as High Park, may have an influence in the spread of Acer platanoides.

## `crossstreet1` & `crossstreet2`
While `streetname` covers the street a tree lies upon, `crossstreet1` & `crossstreet2` tells us what streets (if any) are in close proximity to the tree. This augments `streetname` by narrowing down the general location of the tree to specific intersections. Similarly to `streetname`, small streets intersection with other small streets may may be places where street trees are planted. Depending on the size and configuration of the intersection, there may be more, or less space for trees 

```{r}
#| label: fig-4
#| fig-cap: Acer platanoides distribution across Toronto crossstreets up to September 2024
#| echo: false
#| warning: false
#| message: false

# Gather 10 cross streets with the highest counts of Acer platanoides
crossstreet1_count <- analysis_data %>%
  filter(botanical_name == "Acer platanoides") %>%
  count(crossstreet1, sort = TRUE) %>%
  head(10)

crossstreet1_count |> 
  ggplot(aes(x = n, y = crossstreet1)) +
  geom_col() +
  theme_minimal() +
  labs(x = "Cross Street",
       y = "Acer platanoides count")
```

```{r}
#| label: fig-5
#| fig-cap: Acer platanoides distribution across Toronto crossstreets up to September 2024
#| echo: false
#| warning: false
#| message: false

# Gather 10 cross streets with the highest counts of Acer platanoides
crossstreet2_count <- analysis_data %>%
  filter(botanical_name == "Acer platanoides") %>%
  count(crossstreet2, sort = TRUE) %>%
  head(10)

crossstreet2_count |> 
  ggplot(aes(x = n, y = crossstreet2)) +
  geom_col() +
  theme_minimal() +
  labs(x = "Cross Street",
       y = "Acer platanoides count")
```
Comparing @fig-4 & @fig-5, most street trees are placed well clear of intersections. Of the streets that are located in proximity to an Acer platanoides, Bathurst St and Mount Pleasant Rd stick out. Both are prominent in @fig-4 and @fig-5. Augmenting this with @fig-3, we can see that Bathurst St is seen in all three graphs, suggesting than many of Acer platanoides' trees grow along that corridor, and perhaps this is how they spread across the city.

# Measurement
Urban Forestry Management oversees trees within the City of Toronto. Data used in Street Trees comes from a program called Area Street Tree Maintenance (ASTM) [@StreetTreeMaintenance]. ASTM uses a form of cluster sampling to collect data and preform maintenance on street trees. Forestry Management gets information on a general area, which we can call a cluster. Within that cluster, each street tree is examined, and forestry management makes sure they scour through every neighborhood within the cluste [@StreetTreeMaintenance]r. The Street Tree dataset therefore, is composed of these clusters, and every street tree found within that cluster. The way clusters are determined is through request by citizens [@StreetTreeMaintenance]. This does mean that some clusters may overlap with each other, especially as requests can be made in the exact same location several times. However, trees that are recorded multiple times will stand to be more accurate over time. This is because the dataset is constantly changing. It is not a snapshot of street trees in Toronto at a specific time. It is meant to be an up-to-date representation of street trees in Toronto. Trees that aren't recorded as often may have their records be more out of date. For example, dbh_trunk is a varaible that will slowly increase with time (barring tree removal or extreme modifications otherwise). A record 2 or 3 years behind the current date would display a dbh_trunk value that may not reflect the current value. This may be a reason ASTM is looking towards city-wide periodic checks, with reports suggesting tree pruning every seven years.

`botanical_name` is derived when Forestry Management successfully identifies the species of a tree. The scientific name is recorded in the dataset, along with it's common English name (but this report doesn't use the English name). The location of the tree is then allocated to a postal address of a (private) property. This does not mean the tree is located in this postal address, just that it is in proximity to it. A postal address will have a streetname, recorded as `streetname`, as well as unit number (which isn';'t used in this paper). Given the postal address of a street tree, crosstreets can be determined by map, or by on-site teams overseeing any maintenance at that location. This is how `crossstreet1` and `crossstreet2` are documented. Finally, using a postal address allows for the tree to be located by ward. Boundaries for each ward are listed on the City of Toronto's website, allowing for easy categorization.

# Model

Our modeling strategy estimates street trees by botanical name. To achieve this, I created a linear model using location variables as predictors. The goal is to see if Acer plnatoides has a distinct spread by location

The model created uses multiple linear regression, predicting botanical name (`botanical_name`) asa function of 4 location varaibles, `ward` (city ward), `streetname`(tree street is on/nearby), and `crossstreet1` & `crossstreet2` (nearest intersecting streets, if applicable).

**Linear Model by Ward, Streetname, Crossstreet1, and Crossstreet2**\
\begin{align}  
y_i &= \beta_0 + \beta_1 \cdot \text{ward}_i + \beta_2 \cdot \text{streetname}_i + \beta_3 \cdot \text{crossstreet1}_i + + \beta_4 \cdot \text{crossstreet2}_i + \epsilon_i \\
\epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align}

where:

-   $y_i$ represents the botanical name of the tree $i$.
-   $beta_0$ is the intercept.
-   $beta_1$, $\beta_2$, $\beta_3$ , $\beta_4$ represent the coefficients for the predictors: `ward`, `streetname`, `crosstreet1`, and `crossstreet2`, respectively.
-   $epsilon_i$ is the error term, which is assumed to be normally distributed with variance $\sigma^2$.

This model helps capture the general trend tree species across an array of location factors in Toronto. The model was then used to predict botanical names, which were added to the original datasets as new columns. The augmented dataset allowed for a comprehensive comparison of predicted versus observed botanical names, and facilitated visualization of how widespread certain species are across Toronto.

## Model set-up

To model the species of street trees, we used multiple linear regression to estimate the relationship between tree species (`botanical_name`) and several lcoation predictors, including the ward, street name, and cross street names. This section provides a detailed breakdown of how we constructed the models, including the selection of predictors and the data processing steps taken to ensure that the models capture meaningful patterns.

## Feature Selection

We used the following predictors to model the polling percentage (pct) for both candidates:

**1.City Ward (ward):** - The most general location metric in the dataset. We are looking for relationship between species and location, so starting with 25 sections of the city is a good start. The city ward gives a general location for a street tree in Toronto. Given unique characteristics for each of the wards, we can use those to postulate reasons behind a specific species' prominence

**2.Nearest Street (streetname):** - These next highest location data is the street name. This can help pinpoint specific roads that certain species grow along. Since many roads pass through wards, it can help paint a picture for trends seen in wards data. While wards provide a general idea of location, they come at the cost of precision. Streets can narrow down a location of a tree to a simple line or curve, as opposed to an area. Streets can also cross between wards, which provides potential reasoning for relationships between wards.

**3.Nearest Cross Street(s) (crossstreet1 & crossstreet2):** - Because streets can span across a ward, or multiple wards, knowing where slong the street a tree is located is helpful. Cross streets define a tree's nearest intersection. Streets are often long lines, which is a problem for precision. Referencing cross-streets for our model can help pinpoint where exactly on a street, a tree is located. This brings the precision down to plots of land of a few square meters.

## Model Training and Prediction

-   The model was fitted to predict botanical_name as a function of ward, streetname, crossstreet1 and crossstreet2. This model helps in understanding the general trends in tree species, across the City of Toronto.
-   The model was trained using the lm() function in R, which estimates the parameters by minimizing the sum of squared residuals. 
-   After fitting the model, predictions were made and added to the dataset. This allowed us to compare actual botanical name with those predicted by the models.

## Model Assumptions

The following assumptions underlie the multiple linear regression models used:

-   **Linearity:** We assume that the relationship between the predictors (ward, streetname, crossstreet1, crossstreet2) and the outcome (botanical_name) is linear.
-   **Independence:** Each poll is treated as an independent observation. We assume that polling percentages from different polls are not influenced by one another.
-   **Homoscedasticity:** The variance of the residuals is assumed to be constant across all levels of the independent variables.
-   **Normality of Residuals:** The error term (\epsilon\_i) is assumed to be normally distributed, which is required for the validity of hypothesis testing.
-   **No Multicollinearity:** We checked that the predictors are not highly correlated, to ensure that the coefficients estimated by the models are reliable and that multicollinearity does not bias the results.

These assumptions are crucial for ensuring that the model coefficients are unbiased and that the predictions are meaningful.

## Cross-Validation and Overfitting Prevention

To evaluate the robustness of our models, a train-test split was applied. The training data was used to fit the model, while the test data (which the model had not seen during training) was used to evaluate model performance. This approach helps in identifying over fitting, ensuring that the model can generalize beyond the specific dataset used for training.

For future improvements, we could explore the use of cross-validation for further robustness checks, ensuring that the model’s performance is not overly dependent on a particular train-test split.

### Model justification

The choice of multiple linear regression models for predicting the species of street trees in Toronto is driven by several key considerations regarding the nature of the data and the objectives of the analysis. Below, we provide a detailed justification for the chosen models and predictors, discuss the reasoning behind the modeling choices, and address limitations and alternatives considered during the process.

# Results

```{r}
#| label: fig-6
#| fig-cap: Plot of ward by tree species. The data used is generated from the linear model predicting botanical_name from ward, streetname, crossstreet1, and crossstreet2.
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
model_data <- read_parquet(here::here("data/model_data/model_data.parquet"))
model <- readRDS(here("models/model.rds"))

ggplot(model_data, aes(x = ward)) +
  geom_point(aes(y = botanical_name), color = "black") +
  geom_line(aes(y = fitted_name), color = "blue", linetype = "dotted") +
  theme_classic() +
  labs(y = "Botanical Name", x = "Ward", title = "Linear Model over Ward")
```
@fig-6 shows that across all wards, only some trees are prevalent across all of them. One of them is Acer platanoides.
```{r}
#| label: fig-7
#| fig-cap: Plot of ward by tree species. The data used is generated from the linear model predicting botanical_name from ward, streetname, crossstreet1, and crossstreet2.
#| echo: false
#| warning: false
#| message: false

ggplot(model_data, aes(x = streetname)) +
  geom_point(aes(y = botanical_name), color = "black") +
  geom_line(aes(y = fitted_name), color = "blue", linetype = "dotted") +
  theme_classic() +
  labs(y = "Botanical Name", x = "Street", title = "Linear Model over Street")
```
In @fig-7, Acer platanoides is seen across the most streets, given the frequency of points. 
```{r}
#| label: fig-8
#| fig-cap: Plot of crossstreet1 by tree species. The data used is generated from the linear model predicting botanical_name from ward, streetname, crossstreet1, and crossstreet2.
#| echo: false
#| warning: false
#| message: false

ggplot(model_data, aes(x = crossstreet1)) +
  geom_point(aes(y = botanical_name), color = "black") +
  geom_line(aes(y = fitted_name), color = "blue", linetype = "dotted") +
  theme_classic() +
  labs(y = "Botanical Name", x = "Cross street 1", title = "Linear Model over Cross Street 1")
```
```{r}
#| label: fig-9
#| fig-cap: Plot of crossstreet2 by tree species. The data used is generated from the linear model predicting botanical_name from ward, streetname, crossstreet1, and crossstreet2.
#| echo: false
#| warning: false
#| message: false

ggplot(model_data, aes(x = crossstreet2)) +
  geom_point(aes(y = botanical_name), color = "black") +
  geom_line(aes(y = fitted_name), color = "blue", linetype = "dotted") +
  theme_classic() +
  labs(y = "Botanical Name", x = "Cross street 2", title = "Linear Model over Cross Street 2")
```
Both @fig-8 and @fig-9 carry the same conclusions over to crossstreet1 and crossstreet2
\newpage

# Discussion

The model results don't show a clear direction Acer platanoides travels in. Instead it tells us that it is present in almost every ward, and every street in Toronto. This undermines one of the key issues with this tree, the way it spreads. Its seeds are know to travel long distances, which is how it can travel to other cities and towns. This is exacerbated by the sheer number of these trees in Toronto. All those trees will spread their seeds across Southern Ontario, disrupting the natural treescape. This report has shown us the importance of mintoring and controlling the population of Acer platanoides in Toronto and Southern Ontario.

## Weaknesses and next steps

Street Tree data is cumulative, meaning Urban Forestry has to manually update the status of each tree, which is an arduous task for over 30000 trees. As a result the data may be mismatched, with some trees having only been checked months or even years prior. This may be why @fig-4 is so skewed. There may have been errors in the way the data was inputted for certain trees, leading to misplaced decimals. A similar occurrence happened when observing the raw data, which can be seen in the Appendix. No additional information was provided by OpenData Toronto when viewing the dataset on their database, apart from who provided the data, Urban Forestry. Perhaps more detail on their part, including units of measurement for dbh_trunk, would have helped make this analysis much easier.Another limitation mentioned earlier is the inclusion of some privately-owned trees.

Next steps would be to investigate the high counts of Acer platanoides in Toronto. Where did they originate from? Where did they start showing up and who was responsible for that? From there, a plan could be drafted to prevent the future spread of invasive plants, and give insight to how Acer platanoides can be stopped in Toronto

With newer reports, pressure will be put upon Urban Forestry to better curtail the data they have collected thus far. It is imperative that their hard work be recognized and any feedback on the data they have collected be sent to them as well, ensuring better datasets for analysts, leading to more informative conclusions for all.

\newpage

# Appendix

## Surveys, sampling, and observational data

ASTM's survey of Toronto's street trees leaves a lot to be desired. With their implementation of cluster-sampling, certain "clusters" of the city may be left unrecorded for extended periods of time, creating variance in accuracy. Additionally, there seems to be some quirks with the collected data. According to @fig-9, there are several extreme values for the variable `dbh_trunk` that make seemingly no sense. The tree with the largest recorded dbh comes in at around 800 cm, which would make it one of the girthiest trees in Canada, according to Monumental Trees' records [@MonumentalTrees]. These values seem to be mistakes made by management misinputting the recorded dbh value, or OpenData Staff improperly cleaning the data.

While ASTM is pledging to move forward with periodic city-wide maintenance [@StreetTreeMaintenance], which should include updating records on all trees, more can be done to better preserve street tree data. First, given how the dataset will cahnge over time, a periodic update should be implemented in a street tree survey. However, instead of replacing data outright, new datasets should be created to correspond with each cycle of data collection. Just like ASTM did, we can implement cluster-sampling. Toronto can be split into smaller clusters, ensuring each cluster is distinct from one another. Within each cluster, a team will work to cover all neighborhoods, identifying and then recording street trees in each neighborhood. Each tree will have it's species and common name identified. Then the tree's location will be given by both coordinates (for precision), and approximated to the nearest postal address (for context). The nearest postal address will have separate variables for number and street. Additionally, assuming road junctions in Toronto max out at 4 roads, 3 additional cross streets can be listed as variables, as a way to pinpoint the nearest junction to a street tree. DBH is a relatively vague measurement, as breast height changes from location to location. Therefore this survey will use Brokaw and Thompson's proposed $D_x$ measurement [@dbhalt]. This eliminates the vaugeness of "breast height", and instead outlines the recorded height where the diameter measurement took place. A new varaible to add would be height, as that is a key physical indicator of tree growth. This can be done by taking measurements along the ground and using trigonometry to calcualte tree height, as shown by the University of British Columbia [@bigTree].


```{r}
#| label: fig-10
#| fig-cap: Distribution of Distance-at-Breast-Height diameter of all trees in Toronto up to September 2024
#| echo: false
#| warning: false
#| message: false

analysis_data |> 
  ggplot(aes(x = dbh_trunk)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = "DBH Diameter (cm)",
       y = "Tree Count")
```

## Data Cleaning

While cleaning that dataset, some inconsistent values were discovered. A particular species of tree, Salix x sepulcralis, had it's name misspelled in the dataset. This may have come from using special characters that could not be saved onto the dataset, resulting in a garbled name. An additional step was taken to restore the correct botanical name, using the common name as reference. Similarly, one observation did not have a ward location allocated. This was remedied by reading the global coordinates from geometry, a variable in the raw data file that shares the latitude and longitude of the tree individual. Using these coordinates, and the address number, the general location for this tree was determined, along with its ward location. The ward entry for this observation was updated accordingly.

# References
